The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-22 paper/ms.tex
revisions in introduction

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a revolution in the
era of {\em Gaia} DR2. Recent major discoveries include the remnants of a
major merger \citep{2018ApJ...860L..11K, 2018Natur.563...85H,
2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space ``spiral'' in the
solar neighborhood \citep{2018Natur.561..360A} possibly indicating either
local substructure infall \citep{2018MNRAS.481.1501B, 2018arXiv180800451L} or
bar buckling \citep{2018arXiv181109205K}, and a gap suggestive of perturbation
by a dark matter subtructure in the tidal stream GD1
\citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center is taken to be $\sim 8.1\text{--}8.3\,\kpc$
\citep{2009ApJ...692.1075G,2018AA...615L..15G}, with a recent $\sim0.3\%$ measurement from the GRAVITY collaboration \citep{2019arXiv190405721A}. The Sun's vertical distance
from the Galactic midplane, the other independent measurement that sets this
coordinate system, can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars), and is usually taken to be $\sim 25\,\pc$ \citep{2001ApJ...553..184C},
with more recent measurements from {\em Gaia} DR2 placing it at $\sim 20\,\pc$
\citep{2019MNRAS.482.1417B}. Another strategy is to use the cold gas or HII
regions in the disk to define the Galactic midplane, leading to slightly
different values ($\sim 5\,\pc$) for the Sun's relative height
\citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review of these
measurements is given by \citet{2016ARAA..54..529B}.

Once a Galactocentric coordinate system has been established and a 6D phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities which describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the typical integral of the Hamiltonian in cylindrical coordinates and are
given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We typically write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

>>>>

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a
revolution 
%mm in the era of {\em Gaia} DR2. 
     as a result of {\em Gaia} Data Release 2 (DR2).  Recent major discoveries include the remnants of a
major merger \citep{2018ApJ...860L..11K, 2018Natur.563...85H,
2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space ``spiral'' in the 2019arXiv190405721A
solar neighborhood \citep{2018Natur.561..360A} possibly indicating either
local substructure infall \citep{2018MNRAS.481.1501B, 2018arXiv180800451L} or
bar buckling \citep{2018arXiv181109205K}, and a gap suggestive of perturbation
by a dark matter subtructure in the tidal stream GD1
\citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center
%mm is taken to be $\sim 8.1\text{--}8.3\,\kpc$
    can be measured
\citep{2009ApJ...692.1075G,2018AA...615L..15G}, with a recent 
%mm $\sim0.3\%$ measurement from the GRAVITY collaboration
    measurement using near-infrared interferometry of $8.178 \pm
    0.035$~kpc 
\citep{2019arXiv190405721A}. The Sun's vertical distance
from the Galactic midplane, the other independent measurement that sets this
coordinate system, can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars), and is usually taken to be $\sim 25\,\pc$ \citep{2001ApJ...553..184C},
with more recent measurements from {\em Gaia} DR2 placing it at $\sim 20\,\pc$
\citep{2019MNRAS.482.1417B}. Another strategy is to use the cold gas or HII
regions in the disk to define the Galactic midplane, leading to slightly
different values ($\sim 5\,\pc$) for the Sun's relative height
\citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review of these
measurements is given by \citet{2016ARAA..54..529B}.

Once a Galactocentric coordinate system has been established and a 6D phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities that describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the %mm typical
 integral of the Hamiltonian in cylindrical coordinates and are
given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We typically write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.
